Domain-Independent Detection of Emergency Situations\\Based on Social Activity Related to Geolocations

### 2018-02-23 paper.tex
add introduction to several sections

\author{Sergio Campos}
%\authornote{This author is the
 % one who did all the really hard work.}
\affiliation{%
  \institution{Department of Geophysics, University of Chile}
%  \streetaddress{1 Th{\o}rv{\"a}ld Circle}
  \city{Santiago} 
  \country{Chile}}
\email{jaime@dgf.uchile.cl}

@@

\begin{abstract}
%Most methods to detect emergency situations using Twitter rely on keyword. The problem of keyword-based methods is the need to train in specific domains and language for different type of events, for example: earthquakes, typhoons, terrorist attacks, tornadoes, etc. Our proposal is to rather use the recurring mention of a country locations in microblogging messages to identify such events without using keywords and characterize through of interarrival time probability distributions. Our method use a SVM classifier that is independent of the text language of the messages and can be training and testing among different languages and domains.
%
Most methods for detecting emergency situations using Twitter rely on identifying features within messages that contain domain-specific keywords. However, keyword-based methods require models to be trained on historical data of specific domains, in multiple languages, and for different types of events (e.g., earthquakes, floods, wildfires, etc.). 
%
In addition to being costly, these approaches may fail to detect previously unexpected situations, such as uncommon catastrophes or terrorist attacks.
%
Nevertheless, collective mentions of keywords are not the only type of self-organizing phenomena that may arise when a real-world extreme situation occurs.
%
Just as physical sensors become activated when stimulated, localized citizen sensors will also react in a similar manner.  
%
To leverage this information, we propose to use self-organized geolocation related activity to identify emergency situations.
%
We propose to detect such events by tracking the frequencies, and probability distributions of the interarrival time of the messages related to specific locations.
%
Using an off-the-shelf classifier that is independent of domain-specific features, we study and describe emergency situations based solely on location-based features in messages.
%
Our findings indicate that anomalies in location-related social media user activity indeed provide information for automatically detecting emergency situations independent of their domain.
\end{abstract}

@@

During emergency situations, traditional media may suffer infrastructure issues and real-time communications could be disrupted. Instead, microblogging has played a critical role over the last fifteen years allowing users to share real-time information from people local to the incident, such as status updates, casualties, damages and alerts \cite{kumar2011tweettracker,imran2013extracting,stowe2016identifying,reuterfifteen}. For this reason, researchers have studied user behavior during these events to detect, summarise and classify messages with the goal of helping authorities, and the general public, with situational awareness to provide fast and conscientious responses during crisis situations.

Twitter is a microblogging platform that allows users to share short messages (called \textit{tweets}) and is currently used worldwide by over $300$ million people.\footnote{http://www.statista.com/statistics/282087/number-of-monthly-active-twitter-users/} About $80\%$ of Twitter users access from mobile devices, which contributes to the immediacy of diffusion of information, especially during crisis situations \cite{castillo2016big}. One of the most important tasks during these situations, is to timely detect incoming real-world events. This is because, as with most social media conversations, messages are often overridden with irrelevant and redundant noise. In current works \cite{kumar2011tweettracker,caragea2011classifying,imranaidr2014,maldonado2017}, these tasks are solved with methods that rely on keyword based filters on the \textit{Twitter Public Streaming API}.\footnote{https://developer.twitter.com/en/docs/tweets/filter-realtime/overview} The problem with keyword-based methods is the need to train in specific domains for different type of events. For instance, \citeauthor{olteanu2014} \cite{olteanu2014} generate a set of keywords based on different datasets. However, this is not sufficient for cases in which specific and previously unseen terms arise for a particular event (e.g., \textit{\#eqnz} for Earthquakes in New Zealand, or \textit{\#pabloph} for Typhoon Pablo in Philippines) \cite{potts2011tweeting,bruns2012local,karimi2013classifying}. Furthermore, these sets of keywords are commonly obtained a particular language and will not work in others.

Additionally, several works related to crisis situations show a strong relationship between a type of event (what) and the spatiotemporal dimensions (when and where). For example, it has been observed that top trends during earthquakes are related to location mentions \cite{mendoza2010twitter}, also, that there is a strong relationship between the proximity to a hurricane path and hurricane-related social media activity \cite{kryvasheyeu2016rapid}. Other works have addressed the extraction of locations and points of interest during floods \cite{lingad2013location}, as well as mixing geographic information system (GIS) information with geo-tagged messages to improve disaster mapping and real-time event tracking \cite{huang2015disastermapper}.

Based on previous works, the question we address here is: \textit{Is there evidence to detect an emergency situation based on anomalies in messages related to contain locations?}.

Here, we propose a method based on recurring references to country level locations in message metadata. For this task, we create a gazetteer tree based on the hierarchy of the specific country, divide the messages into fixed time-windows and compute the frequency and the probability distributions of the interarrival time of the messages for each geographical hierarchy. To detect an emergency situation, we train a SVM classifier by each hierarchy and apply a geographic spread to filter false positives detection considering that an emergency situation can be \textit{Focalized} or \textit{Diffused}.

Our main contribution is to create a methodology that detects instantaneous emergency situations just by using the location related information for a specific country. Furthermore, our classifiers do not depend on language to detect a new event since they do not use textual features as input. Moreover, we characterize crisis situations that affect small or large geographical areas.

The paper is organized as follows: we first introduce an overview of relevant literature related to emergency situation and event detection in social media. Next we present a complete description of our proposal divide in four parts. Thereafter, we provide a full description of our dataset and later we summarise our experimental validation over the ground truth and on-line evaluation. Finally, we deliver our discussions, conclusions and future work.

@@

In addition to crisis-related social media monitoring, there are also some unsupervised event detection approaches based on location information. \textit{Jasmine} \cite{watanabe2011jasmine} detected local real-world events using geolocation information from microblog documents. To detect such events, they identify a group of tweets that describe a particular theme, which are generated within a short time frame and a same geographic area. Similarly, \citeauthor*{unankard2015emerging} \cite{unankard2015emerging} proposed an approach for early detection of emerging hotspot events in social networks with location sensivity. In this work, the authors identified strong correlations between user locations and event locations when detecting emerging events using content similarity between clusters. In the work of \citeauthor{walther2013geo} \cite{walther2013geo} real-world events were detected in a small scale with messages from the New York metropolitan area. There, clusters were created for each candidate event and evaluated using cluster score based on textual features (sentiment analysis, common theme, duplicate, etc) and other features (tweet count, unique coordinates, etc.).

Supervised approaches have also been proposed. \textit{TEDAS} \cite{li2012tedas} detected, analysed and identified events using refined rules (e.g., keywords, hashtags) and classify messages based on content as well Twitter specific features as URLs, hashtags and mentions. Besides, location information is extracted using both explicit geographical coordinates and implicit geographical references in the content. In the same way, \citeauthor{becker2011beyond} \cite{becker2011beyond} proposed an on-line clustering technique, which continuously clusters similar tweets and then classifies the clusters using Support Vector Machine algorithm. Finally, events (clusters) are classified into real-world events or nonevents.

@@

\noindent{\textbf{Summary.}} Most previous works to detect an emergency situations rely on keywords using probabilistic temporal models for specific domains (e.g., keywords or location). In general terms, these approaches need knowledge about the event and a new emerging topic referring to a crisis cannot be identified to detect an emergency situation. In the case of the works on event detection based on locations, they are designed to be applied in a small area and based on historical data. Regardless of which approach is implemented (supervised or unsupervised), textual features are the most used attributes to characterise an event and on-line clustering is the most common technique to create candidate events.

Based on the works presented by \citeauthor{guzman2013line} \cite{guzman2013line} and \citeauthor{maldonado2017} \cite{maldonado2017}, we extend the ideas of bursty keywords and \textit{z-score} variation between fixed time-windows and we apply these proposals over locations to identify anomalies in social media activity. Furthermore, we do not use a set of keywords to detect specific types of events (e.g., earthquake, flood or terrorist attack). In contrast, our proposal is focused on detecting such events by tracking frequencies, and probability distributions of the interarrival time of the messages related to specific locations.

@@

In order to provide a complete coverage of location-based detection of emergency situations, we propose an approach which has four stages (depicted in the Data Process module in Figure \ref{fig:diagram}). Next, we describe each stage:

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{img/diagram2.png}
	\caption{Key components of the proposed approach.}
	\label{fig:diagram}
\end{figure}

@@

%To increase the number of messages related to locations, we consider tweets and retweets messages. 
Since we consider users as citizen sensors, we filter messages depending of native language of each country using the attribute \textit{lang} retrieved from the metadata in tweets.\footnote{https://developer.twitter.com/en/docs/tweets/data-dictionary/overview/tweet-object} For example, if we analyse Italy, we just consider messages in Italian. Finally, we remove user mentions, URLs,  special characters and apply text tokenization. Also we do not remove hashtags or stopwords because some locations can be included in hashtags terms and some places can contain stopwords that can differentiate them from other similar non-related to locations terms.

\subsection{Signal Creation}
The problems that we address in this stage are how to infer locations from microblog messages to track signals over time in the data stream and what is the lowest hierarchy level to find and assign to these signals.

\subsubsection{Geographical Hierarchy}\label{sssec:geohie}
Using the idea of \textit{Gazetteer as a Tree} presented in \cite{yin2014pinpointing} in which each place is associated with a canonical taxonomy node, we construct our gazetteer tree based on Geonames\footnote{http://download.geonames.org/export/dump/} and Wikipedia\footnote{http://www.wikipedia.org/} for each country to analyse. However, in \cite{yin2014pinpointing} the gazetteer hierarchy presents four levels where the lowest level represents a specific point of interest. In our approach, we use a subset of the gazetteer hierarchy with three levels, in which the lowest level is represented by a city since a large amount of users specify their location down to this level \cite{hecht2011tweets}. For example, if we have the \textit{city:Manchester}, we associate this location with \textit{region-state:North West} and also with \textit{country:England}. Furthermore, the name of locations are considered just in the native language of the country. For instance, in the case of Italy locations, we consider \textit{Roma} and not \textit{Rome}.

\subsubsection{Finding Locations in Microblog Metadata}

The structure of the Tweet metadata obtains information about the message and the user who shares a message. Considering the geographical hierarchy explained in the previous section, we search these locations on different levels of metadata and create one signal for each as following:

\begin{itemize}
	\item Tweet Text: location is mentioned in the attribute \texttt{text} of tweet object, that is, on the body of the message.
	\item User Location: location is mentioned in the attribute \texttt{location} inside the \texttt{user object}, that is, the location set by the user in his profile.
	\item Tweet Text - User Location: location is mentioned in the attribute \texttt{text} of tweet object and also location is mentioned in the attribute \texttt{location} inside the \texttt{user object}. This means that the location is mentioned in the body of the message and the user who shares message has the same location in his profile. In this case, tweet text and user location can be different in the smallest hierarchy, but in the highest level can be equal. 
\end{itemize}

In this way, mixing geographical hierarchy and locations in microblog metadata, we create $NxM$ signals where $N$ is the number of locations obtained by gazetteer tree and $M$ is the number of metadata-levels extracted from Tweet Object. For instance, we create a signal for \textit{city:Manchester} and we find this hierarchy in \textit{metadata:Tweet Text} and also in \textit{metadata:User Location}. That means that we track the mention of \textit{city:Manchester} at the level of the body of message and at the level of the location of the user profile individually.

\subsection{Time-Windows}
In this stage we address the problems of how to divide and determine the time-window size to detect a new emergency situation and what features by the time-window allow it.

@@

According to \citeauthor{guzman2013line} \cite{guzman2013line}: ``If the window size is too small, the ocurrence of empty windows for a term increases, making the noise rate increase and frequency rate tend towards zero. On the other hand, if the window size is too large, the stability of the signal becomes constant and bursty keyword detection is delayed''. Using this definition, we divide our signals into windows of six minutes because it divides a 24-hour day exactly, making the analysis easier to understand and to compare from different days.

@@

To characterize the urgency of the messages during a time-window, we use our previous results \cite{DBLP:conf/fdia/Sarmiento17} to compute the \textit{interarrival time} which is defined as $d_{i} = t_{i+1} - t_{i}$, where $d_{i}$ denotes the difference between two consecutive social media messages $i$ and $i+1$ that arrived in moments $t_{i}$ and $t_{i+1}$ respectively. Using this definition and according the work of \citeauthor{kalyanam2016prediction} \cite{kalyanam2016prediction}, high-activity events have a high-frequency in the first bins represented by values $d_{i} \approx 0$.

@@

%\subsection{Subsetting Signals}
%\begin{itemize}
%	\item Several signals have a null values along the time. For example, any city that is not a big city or capital city, does not a continuous frequency. To avoid this behaviour, we use cities just for compute the state frequency signal and later we discard them. Finally, we just monitor states and country hierarchy.
%	\item Furthermore, we remove gps signal because just 1-3 percent of messages have gps activated.
%\end{itemize}

@@

An emergency situation that affects and mobilizes response in a small area is defined as \textit{focalized}, while a disaster with a large geographic impact is defined as \textit{diffused} \cite{olteanu2015expect}. Using this definition, we extend this concept to represent neighborhoods between locations obtained from section \ref{sssec:geohie}. For that purpose, we create an \textit{adjancency matrix} $M$, where $M_{i,j} = 1$ represents if two locations are geographically connected and $M_{i,j} = 0$ if they are not connected. For instance, if an event is diffused (e.g., earthquake), the detection should be in adjacent-locations independentely of metadata-level. On the other hand, if an event is focalized (e.g., terrorist attack), just one location should be detected but in different metadata-levels simultaneously.

@@

\section{Dataset}

Data is collected from Twitter Public Streaming API, which allows access to subsets equal to $1\%$ of public status descriptions in real-time. With this tool, we can retrieve either messages using a set of keywords or messages from specific locations setting a bounding box. In our approach, we get entire subsets of messages without use keywords or specific locations. Then, we retrieve random messages about any topic and any place in the world.

@@

We analyse five earthquakes with magnitudes between $5.5Mw$ and $7.6Mw$\footnote{Mw: the moment magnitude scale}, ocurring in Italian-speaking and Spanish-speaking countries between October 2016 and April 2017 (Table \ref{tab:eqs}). For that purpose, we collect $20$ million of messages 12-hours before and after the emergency situation events.

@@

\section{Experiments}

@@

To classify messages we employed traditional binary classifier Support Vector Machine \textit{(SVM)}. As a result of the analysed data scattering (Figure \ref{fig:scatter}), we separated country and state in different datasets and set both kernels and classification parameters independentely. On the one hand, \textit{country} classifier uses a polynomial kernel and strict-parameters for gamma, cost and weights since a great amount of messages are included in country hierarchy as an effect of the minor hierarchies. On the other hand, \textit{state/region} classifier uses a linear kernel with soft-weights and cost.

@@

Given that an emergency situation is not an usual event, we have an highly unbalanced data respect to the classes after labeled ($1\approx2\%$ of positive class corresponding to \textit{detection}). Therefore, we used \textit{under-sampling} \cite{lunardon2014rose} over country and state datasets increasing our possitive class to $15\approx18\%$. Additionally, to validate our model, we used \textit{5-fold cross-validation} where one earthquake dataset is used as testing and the remaining earthquakes dataset as training.

@@

In addition to the analysis of number of detections by labels, we also study the number of detections by time-windows. For this analysis, we search the time-windows for each hierarchy where the all metadata-levels are well classified with correct class. According to the results shown on the middle of the Table \ref{tab:perfomance}, when we analyse country and state independentely the values of Precision, F1 and FPR have worst values than the analysis by label . 
\begin{table}
	\caption{Average performance of 5-fold cross-validation by hierarchy and geographic spread (G.S.)}.
	\label{tab:perfomance}
	\centering
	\begin{tabular}{c|lcccc}
		\toprule
		&Hierarchy &P &R &F1 &FPR\\
		\midrule
		\parbox[t]{2mm}{\vspace{-0.2cm}\multirow{3}{*}{\rotatebox[origin=c]{90}{\small{label}}}} 
		& Country & 0.3 & 0.83 & 0.45& 0.14\\		
		& State &0.35& 0.83& 0.5 & 0.08\\
		\midrule
		\parbox[t]{2mm}{\vspace{0cm}\multirow{5}{*}{\rotatebox[origin=c]{90}{\small{time-window}}}}
		& Country & 0.15& 0.77 & 0.25 & 0.15\\
		& State & 0.17 & 0.88 & 0.29& 0.12\\
		& Country-State& 0.35 & 0.7 & 0.47 & 0.03\\
		& Country(2)-State with G.S. & 1 & 0.64 & 0.78 & 0 \\
		& Country(3)-State with G.S. & 1 & 0.47 & 0.64 & 0\\
		\bottomrule
	\end{tabular}
\end{table}

@@

Considering the geographic spread by states and the number of metadata-levels by country hierarchy, we analysed the results shown on the bottom of the Table \ref{tab:perfomance}. On the one hand, the row with value equal to \textit{Country(2)-State with G.S.} represents the detection when we considered at least two metadata-levels for the country hierarchy and the geographic spread for states. In contrast to the the previous analyses, we improved the values of the Precision, F1 and FPR. The last metrics are very important because there are no time-windows incorrectly assigned as emergency situations. Consequently, the Recall values decrease which means that our method remove some time-windows classified as detection. Beside the percent of emergency situations detected is equal to $100\%$ with a average delay equal to $10.4$ minutes ($min=6$, $max=14$) from the impact of the event to the first detection.

@@

>>>>

\author{Jaime Campos}
%\authornote{This author is the
 % one who did all the really hard work.}
\affiliation{%
  \institution{Department of Geophysics, University of Chile}
%  \streetaddress{1 Th{\o}rv{\"a}ld Circle}
  \city{Santiago} 
  \country{Chile}}
\email{jaime@dgf.uchile.cl}

@@

\begin{abstract}
%Most methods to detect emergency situations using Twitter rely on keyword. The problem of keyword-based methods is the need to train in specific domains and language for different type of events, for example: earthquakes, typhoons, terrorist attacks, tornadoes, etc. Our proposal is to rather use the recurring mention of a country locations in microblogging messages to identify such events without using keywords and characterize through of interarrival time probability distributions. Our method use a SVM classifier that is independent of the text language of the messages and can be training and testing among different languages and domains.
%
In general, existing methods for automatically detecting emergency situations using Twitter rely on features based on domain-specific keywords found in messages. This type of keyword-based methods usually require training on domain-specific labeled data, using multiple languages, and for different types of events (e.g., earthquakes, floods, wildfires, etc.). 
%
In addition to being costly, these approaches may fail to detect previously unexpected situations, such as uncommon catastrophes or terrorist attacks.
%
However, collective mentions of certain keywords are not the only type of self-organizing phenomena that may arise in social media when a real-world extreme situation occurs.
%
Just as nearby physical sensors become activated when stimulated, localized citizen sensors (i.e., users) will also react in a similar manner.  
%
To leverage this information, we propose to use self-organized activity related to geolocations to identify emergency situations.
%
We propose to detect such events by tracking the frequencies, and probability distributions of the interarrival time of the messages related to specific locations.
%
Using an off-the-shelf classifier that is independent of domain-specific features, we study and describe emergency situations based solely on location-based features in messages.
%
Our findings indicate that anomalies in location-related social media user activity indeed provide information for automatically detecting emergency situations independent of their domain.
\end{abstract}

@@

During emergency situations, traditional media may suffer infrastructure issues and real-time communications could be disrupted. Instead, microblogging has played a critical role over the last fifteen years allowing users to share real-time information from people local to the incident, such as status updates, casualties, damages and alerts \cite{kumar2011tweettracker,imran2013extracting,stowe2016identifying,reuterfifteen}. For this reason, researchers have studied user behavior during these events to detect, summarize and classify messages with the goal of helping authorities, and the general public, with situational awareness to provide fast and conscientious responses during crisis situations.

Twitter is a microblogging platform that allows users to share short messages (called \textit{tweets}) and is currently used worldwide by over $300$ million people.\footnote{http://www.statista.com/statistics/282087/number-of-monthly-active-twitter-users/} About $80\%$ of Twitter users access from mobile devices, which contributes to the immediacy of diffusion of information, especially during crisis situations \cite{castillo2016big}. One of the most important tasks during these situations, is to timely detect incoming real-world events. This is because, as with most social media conversations, messages are often overridden with irrelevant and redundant noise. In current works \cite{kumar2011tweettracker,caragea2011classifying,imranaidr2014,maldonado2017}, these tasks are solved with methods that rely on keyword based filters on the \textit{Twitter Public Streaming API}.\footnote{https://developer.twitter.com/en/docs/tweets/filter-realtime/overview} However, a drawback of keyword-based methods is that they require to train for each specific domain, and for different type of events. For instance, \citeauthor{olteanu2014} \cite{olteanu2014} generated a set of keywords for event detection based on different datasets. However, these keywords are not sufficient for cases in which specific, and previously unseen, terms arise for particular events (e.g., the hashtags: \textit{\#eqnz}, used for Earthquakes in New Zealand, and \textit{\#pabloph}, used for Typhoon Pablo in Philippines) \cite{potts2011tweeting,bruns2012local,karimi2013classifying}. Furthermore, domain-specific keywords are commonly extracted using a particular language and do not apply to others.

In addition, prior work related to crisis situations in social media, show a strong relationship between the type of event (the {\em what}) and spatio-temporal dimensions (the {\em when} and {\em where}). For example, the top-trends during earthquakes are usually related to location mentions \cite{mendoza2010twitter}, and that there is a strong relationship between the proximity to a hurricane path and hurricane-related social media activity \cite{kryvasheyeu2016rapid}. Other works have addressed the extraction of locations and points of interest during floods \cite{lingad2013location}, as well as mixing geographic information system (GIS) information with geo-tagged messages to improve disaster mapping and real-time event tracking \cite{huang2015disastermapper}.

In order to complement prior findings, we propose to address the following research question: \textit{Can we detect automatically emergency situations, based solely on location-related information found in user messages?}.

We propose a method based on recurring references to country-level locations in message metadata. For this task, we create a gazetteer tree based on the hierarchy of the specific country, divide the messages into fixed time-windows and compute the frequency and the probability distributions of the interarrival time of the messages for each geographical hierarchy. To detect an emergency situation, we train a SVM classifier in each hierarchy and apply a geographic spread to filter false positives detection considering that an emergency situation can be either \textit{focalized} (i.e., that affects a reduced area) or \textit{diffused} (i.e., that affects a large area).

Our main contribution is to create a methodology that detects instantaneous emergency situations just by using the location related information for a specific country. Furthermore, our approach does not depend on language to detect a new event since it does not use textual features as input. In addition, we characterize focalized and diffused crisis situations.

The paper is organized as follows: we first introduce an overview of relevant literature related to emergency situation and event detection in social media. Next we present a complete description of our proposal. Then, we provide a full description of our dataset and later we summarize our experimental validation over the ground truth and on-line evaluation. Finally, we deliver our discussions, conclusions and future work.

@@

In addition to crisis-related social media monitoring, there are also some unsupervised event detection approaches based on location information. \textit{Jasmine} \cite{watanabe2011jasmine} detected local real-world events using geolocation information from microblog documents. To detect such events, they identify a group of tweets that describe a particular theme, which are generated within a short time frame and a same geographic area. Similarly, \citeauthor*{unankard2015emerging} \cite{unankard2015emerging} proposed an approach for early detection of emerging hotspot events in social networks with location sensitivity. In this work, the authors identified strong correlations between user locations and event locations when detecting emerging events using content similarity between clusters. In the work of \citeauthor{walther2013geo} \cite{walther2013geo} real-world events were detected in a small scale with messages from the New York metropolitan area. There, clusters were created for each candidate event and evaluated using cluster score based on textual features (sentiment analysis, common theme, duplicate, etc) and other features (tweet count, unique coordinates, etc.).

Supervised approaches have also been proposed. \textit{TEDAS} \cite{li2012tedas} detected, analyzed and identified events using refined rules (e.g., keywords, hashtags) and classify messages based on content as well Twitter specific features as URLs, hashtags and mentions. Besides, location information is extracted using both explicit geographical coordinates and implicit geographical references in the content. In the same way, \citeauthor{becker2011beyond} \cite{becker2011beyond} proposed an on-line clustering technique, which continuously clusters similar tweets and then classifies the clusters using Support Vector Machine algorithm. Finally, events (clusters) are classified into real-world events or nonevents.

@@

\noindent{\textbf{Summary.}} Most of the existing work for detecting emergency situations, rely on keywords using probabilistic temporal models for specific domains (e.g., using keywords or locations). In general, these approaches require background knowledge about the event, and therefore, they will not identify new types of emerging crisis situations. In the case of existing approaches for event detection based on locations, these have been designed for small areas and are based on historical data. Regardless of the type of approach used (supervised or unsupervised), events are always characterized based on textual features, and on-line clustering is the most common technique used to create candidate events.

Based on the works presented by \citeauthor{guzman2013line} \cite{guzman2013line} and \citeauthor{maldonado2017} \cite{maldonado2017}, we extend the ideas of bursty keywords and \textit{z-score} variation between fixed time-windows and we apply these proposals over locations to identify anomalies in social media activity. Furthermore, we do not use a set of keywords to detect specific types of events. In contrast, our proposal is focused on detecting bursty activity related to such events by tracking frequencies, and probability distributions of the interarrival time of the messages in specific locations.

@@

Our focus in this proposal is to detect an emergency situation based on identifying anomalies in social media activity related to locations. In this way, the main task is to extract locations from messages by reducing the noise and irrelevant information. For this reason, data is pre-processed to allow a better analysis.

For location extraction task, we used a geographical dictionary (also known as gazetteer) to create a geographic hierarchy for a specific country. To understand the effect of an emergency situation inside to the country, we created signals in different levels of the geographic hierarchy and the tweet metadata. Afterwards, signals are divided into fixed time-windows and non textual features are computed for each of these. Finally and with the goals of filtering false positives detection, we created a geographic spread based on the proximity between locations by using an adjacency matrix.

In order to provide a complete coverage of location-based detection of emergency situations, we divided our approach into four stages (depicted in the ``data process'' module in Figure \ref{fig:diagram}). Next, we describe each stage:

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{img/diagram4.png}
	\caption{Key components of the proposed approach.}
	\label{fig:diagram}
\end{figure}

@@

Our focus is on localized bursty user activity. Therefore, initially, we filter messages according to the most common language used in each country (using the attribute \textit{lang} in tweet metadata\footnote{https://developer.twitter.com/en/docs/tweets/data-dictionary/overview/tweet-object}). For example, when analyzing Italy, we only take into account messages written in Italian. This helps to filter noise of unrelated messages. We also remove user mentions, URLs,  special characters, and apply text tokenization. We do not remove hashtags nor stopwords, because some locations can be included as hashtags, and some location names contain stopwords, which differentiate them from other locations or other terms.

\subsection{Signal Creation}
We create a set of discrete-time signals for each location, which indicates each time that a message related to a specific location was posted. In order to explain the effect of an emergency situation in a local and national scope, we use the lowest possible geographical hierarchy level available with the aim of comparing the impact in the highest level. Furthermore, we study the anomalies in different metadata levels to understand how locations are shared in Twitter, for instance, either based in the locations setted by users in their profile or sharing location in their messages.

%We use the lowest possible geographical hierarchy level available. 
% described next.
%The problems that we address in this stage are how to infer locations from microblog messages to track signals over time in the data stream and what is the lowest hierarchy level to find and assign to these signals.

\subsubsection{Geographical Hierarchy}\label{sssec:geohie}
We use the idea of \textit{gazetteer as a tree} presented in \cite{yin2014pinpointing} in which each place is associated with a canonical taxonomy node. We create our gazetteer tree based on Geonames\footnote{http://download.geonames.org/export/dump/} and Wikipedia\footnote{http://www.wikipedia.org/}. However, in \cite{yin2014pinpointing} the gazetteer hierarchy presents four levels where the lowest level represents a specific point of interest. In our approach, we use a subset of the gazetteer hierarchy with only three levels: {\em city}, {\em state} and {\em country}. We do so because a large amount of users specify their location down to city level \cite{hecht2011tweets}. For example, if we have the \textit{city:Manchester}, we associate this location with \textit{region-state:North West} and also with \textit{country:England}. As indicated in our data pre-processing stage, we consider only locations in the native language of the country. For instance, in the case of Italy locations, we consider \textit{Roma} and not \textit{Rome}.

\subsubsection{Location Extraction}

The structure of the tweet metadata contains information about the message and the user. Given a small portion of users sharing their current location using GPS coordinates \cite{graham2014world}, we do not consider this level of the tweet metadata in this work.

Considering the aforementioned geographical hierarchy, we extract locations from different parts of the metadata, creating 3 signals for each location:

\begin{itemize}
	\item {\it tweet text:} the location is mentioned in the attribute \texttt{text} of the tweet, that is, on the body of the message.
	\item {\it user location:} the location is mentioned in the attribute \texttt{location} inside the \texttt{user object}, that is, the location set by the user in their profile.
	\item {\it tweet text - user location:} the location is mentioned in the attribute \texttt{text} of tweet object and also location is mentioned in the attribute \texttt{location} inside the \texttt{user object}. This means that the location is mentioned in the body of the message and the user who shares message has the same location in his profile. In this case, tweet text and user location can be different in the smallest hierarchy, but in the highest level can be equal. 
\end{itemize}

In this way, mixing geographical hierarchy and locations in microblog metadata, we create $NxM$ signals where $N$ is the number of locations obtained by gazetteer tree and $M$ is the number of metadata-levels extracted from the tweet object. For instance, we create a signal for \textit{city:Manchester} and we find this hierarchy in \textit{metadata:Tweet text} and also in \textit{metadata:User location}. That means that we track the mention of \textit{city:Manchester} at the level of the body of message and at the level of the location of the user profile individually.

\subsection{Time-Window}
In this stage we address the problems of how to divide and determine the time-window size to detect a new emergency situation and what features by the time-window allow it.

@@

According to \citeauthor{guzman2013line} \cite{guzman2013line}: ``If the window size is too small, the occurrence of empty windows for a term increases, making the noise rate increase and frequency rate tend towards zero. On the other hand, if the window size is too large, the stability of the signal becomes constant and bursty keyword detection is delayed''. Using this definition, we divide our signals into windows of six minutes because it divides a 24-hour day exactly, making the analysis easier to understand and to compare from different days.

@@

To characterize the urgency of the messages during a time-window, we compute the \textit{interarrival time} which is defined as $d_{i} = t_{i+1} - t_{i}$, where $d_{i}$ denotes the difference between two consecutive social media messages $i$ and $i+1$ that arrived in moments $t_{i}$ and $t_{i+1}$ respectively. Using this definition, which follows the work of \citeauthor{kalyanam2016prediction} \cite{kalyanam2016prediction}, high-activity events have a high-frequency in the first bins represented by values $d_{i} \approx 0$.

@@

@@

An emergency situation that affects and mobilizes response in a small area is defined as \textit{focalized}, while a disaster with a large geographic impact is defined as \textit{diffused} \cite{olteanu2015expect}. Using this definition, we extend this concept to represent neighborhoods between locations obtained from section \ref{sssec:geohie}. For that purpose, we create an \textit{adjacency matrix} $M$, where $M_{i,j} = 1$ represents if two locations are geographically connected and $M_{i,j} = 0$ if they are not connected. For instance, if an event is diffused (e.g., earthquake), the detection should be in adjacent-locations independently of metadata-level. On the other hand, if an event is focalized (e.g., terrorist attack), just one location should be detected but in different metadata-levels simultaneously.

\section{Experimental Analysis}

Our experiments aim to find empirical evidences that allow detecting an emergency situation. Thereby, we use the locations for a specific country in different levels of the geographic hierarchy and the tweet metadata. We construct our ground truth based on two publicly available earthquake catalogs and apply the proposal presented in the above section. Thereafter, we study the performance of our methodology mainly based on the concept of affected adjacent locations during an emergency situations.

\subsection{Dataset Description}

We collect data from Twitter Public Streaming API, which allows access to subsets equal to $1\%$ of public status descriptions in real-time. With this tool, we can retrieve either messages using a set of keywords or messages from specific locations setting a bounding box. In our approach, we get entire subsets of messages without use keywords or specific locations. Then, we retrieve random messages about any topic and any place in the world.

@@

@@

We analyze five earthquakes with magnitudes between $5.5Mw$ and $7.6Mw$\footnote{Mw: the moment magnitude scale}, occurring in Italian-speaking and Spanish-speaking countries between October 2016 and April 2017 (Table \ref{tab:eqs}). For that purpose, we collect $20$ million of messages 12-hours before and after the emergency situation events.

@@

\subsection{Methodology}\label{sec:methodology}

According to the ``data classification'' module (Figure \ref{fig:diagram}), we first train a classifier to identify emergency events. Also, we introduce the hierarchy dependence to understand the local and national impact when an high-impact real-world event occurs. Besides, diffused and focalized events are identified with the goal of filtering false positives detections.

@@

To classify messages we employed traditional binary classifier Support Vector Machine \textit{(SVM)}. As a result of the analyzed data scattering (Figure \ref{fig:scatter}), we separated country and state in different datasets and set both kernels and classification parameters independently. On the one hand, \textit{country} classifier uses a polynomial kernel and strict-parameters for gamma, cost and weights since a great amount of messages are included in country hierarchy as an effect of the minor hierarchies. On the other hand, \textit{state/region} classifier uses a linear kernel with soft-weights and cost.

@@

Given that an emergency situation is not an usual event, we have an highly unbalanced data respect to the classes after labeled ($1\approx2\%$ of positive class corresponding to \textit{detection}). Therefore, we used \textit{under-sampling} \cite{lunardon2014rose} over country and state datasets increasing our positive class to $15\approx18\%$. Additionally, to validate our model, we used \textit{5-fold cross-validation} where one earthquake dataset is used as testing and the remaining earthquakes dataset as training.

@@

In addition to the analysis of number of detections by labels, we also study the number of detections by time-windows. For this analysis, we search the time-windows for each hierarchy where the all metadata-levels are well classified with correct class. According to the results shown on the middle of the Table \ref{tab:perfomance}, when we analyze country and state independently the values of Precision, F1 and FPR have worst values than the analysis by label . 
\begin{table}
	\caption{Average performance of 5-fold cross-validation by hierarchy and geographic spread (G.S.)}.
	\label{tab:perfomance}
	\centering
	\begin{tabular}{c|lcccc}
		\toprule
		&Hierarchy &P &R &F1 &FPR\\
		\midrule
		\parbox[t]{2mm}{\vspace{-0.2cm}\multirow{3}{*}{\rotatebox[origin=c]{90}{\small{label}}}} 
		& Country & 0.3 & 0.83 & 0.45& 0.14\\		
		& State &0.35& 0.83& 0.5 & 0.08\\
		\midrule
		\parbox[t]{2mm}{\vspace{0cm}\multirow{5}{*}{\rotatebox[origin=c]{90}{\small{time-window}}}}
		& Country & 0.15& 0.77 & 0.25 & 0.15\\
		& State & 0.17 & 0.88 & 0.29& 0.12\\
		& Country-State& 0.35 & 0.7 & 0.47 & 0.03\\
		& Country(2)-State with G.S. & 1 & 0.64 & 0.78 & 0 \\
		& Country(3)-State with G.S. & 1 & 0.47 & 0.64 & 0\\
		\bottomrule
	\end{tabular}
\end{table}

@@

Considering the geographic spread by states and the number of metadata-levels by country hierarchy, we analysed the results shown on the bottom of the Table \ref{tab:perfomance}. On the one hand, the row with value equal to \textit{Country(2)-State with G.S.} represents the detection when we considered at least two metadata-levels for the country hierarchy and the geographic spread for states. In contrast to the the previous analyses, we improved the values of the Precision, F1 and FPR. The last metric is very important because there are no time-windows incorrectly assigned as emergency situations. Consequently, the Recall values decrease which means that our method remove some time-windows classified as detection. Beside the percent of emergency situations detected is equal to $100\%$ with a average delay equal to $10.4$ minutes ($min=6$, $max=14$) from the impact of the event to the first detection.

@@

\section{Results}

In this section we present an evaluation of our methodology for emergency situation detection. We evaluate our method using different events occurred in England related to terrorist attacks and non emergency situations. In this way, our main goal is to detect emergency events in non labeled data and reduce (as much as possible) the false positives detections. For these tasks, we use the analysis with the best performance presented in the Section \ref{sec:methodology}.
